---
title: "Active Promoters"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath('..'))
knitr::opts_chunk$set(
	error = FALSE,
	message = FALSE,
	warning = FALSE)
```

Loading packages:

```{r packages}
library(tidyverse)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(bedtoolsr)
```

```{r functions, include=FALSE}
select <- dplyr::select
```

# FANTOM5

Data was retrieved from the FANTOM5 project.

```{r data}
TSS <- GRanges(import.bed("data/FANTOM5/TSS_human.bed.gz"))

normal_ctss <- import.bed("data/FANTOM5/breast%2c%20adult%2c%20donor1.CNhs11792.10080-102A8.hg19.ctss.bed.gz")

MCF7_ctss <- import.bed("data/FANTOM5/breast%20carcinoma%20cell%20line%3aMCF7.CNhs11943.10482-107A5.hg19.ctss.bed.gz")

MDA_MB_453_ctss <- import.bed("data/FANTOM5/breast%20carcinoma%20cell%20line%3aMDA-MB-453.CNhs10736.10419-106C5.hg19.ctss.bed.gz")
```

```{r, include=FALSE}
# TSS <- Granges(import.bed("https://fantom.gsc.riken.jp/5/datafiles/phase1.3/extra/TSS_classifier/TSS_human.bed.gz"))
# normal_ctss <- import.bed("https://fantom.gsc.riken.jp/5/datafiles/latest/basic/human.tissue.hCAGE/breast%252c%2520adult%252c%2520donor1.CNhs11792.10080-102A8.hg19.ctss.bed.gz")
# MCF7_ctss <- import.bed("https://fantom.gsc.riken.jp/5/datafiles/latest/basic/human.cell_line.hCAGE/breast%2520carcinoma%2520cell%2520line%253aMCF7.CNhs11943.10482-107A5.hg19.ctss.bed.gz")
# MDA_MB_453_ctss <- import.bed("https://fantom.gsc.riken.jp/5/datafiles/latest/basic/human.cell_line.hCAGE/breast%2520carcinoma%2520cell%2520line%253aMDA-MB-453.CNhs10736.10419-106C5.hg19.ctss.bed.gz")
```

Annotated promoters were retrieved from UCSC, considering 100 nucleotides upstream and 100 nucleotides downstream from the TSS as core promoters.

```{r txdb}
txdb <- promoters(TxDb.Hsapiens.UCSC.hg19.knownGene, upstream = 100, downstream = 100, use.names = FALSE)

data.frame(mcols(txdb)) %>% 
  select(name = tx_name) %>% 
  left_join(read_tsv("data/UCSC/hg19.txt", skip = 1, col_names = c("name", "gene"))) -> mcols(txdb)

data.frame(txdb) %>% 
  select(-name, name = gene) %>% 
  unique() %>% 
  GRanges() %>% 
  export.bed("tables/txdb.bed")
```

According to the [TSS classifier](https://fantom.gsc.riken.jp/5/datafiles/phase1.3/extra/TSS_classifier/TSSpredictionREADME.pdf), TSS were filtered with a strict threshold of *score > 0.228* in order to obtain *true_TSS*.

```{r true_TSS, eval=FALSE}
data.frame(
  name = TSS$name,
  gene = TSS$name,
  score = sub(".*,", "", TSS$name)
) %>% separate_rows(gene, sep = ",") %>% 
  filter(str_detect(gene, "@")) %>% 
  mutate(gene = sub(".*@", "", gene)) -> TSS_scores

strict_TSS <- filter(TSS_scores, score > 0.228)

true_TSS <- subset(TSS, name %in% strict_TSS$name)

data.frame(mcols(true_TSS)) %>% 
  left_join(TSS_scores, by = "name") %>% 
  select(name, gene) %>% 
  distinct(name, .keep_all = TRUE) -> mcols(true_TSS)

write_rds(true_TSS, "tables/true_TSS.rds")
```

```{r, include=FALSE}
true_TSS <- read_rds("tables/true_TSS.rds")
```

The Venn diagram shows promoters across databases. A subset of *true_TSS* not overlapping with annotated promoters was selected as a complement to these regions.

```{r promoters venn_diagram}
intersect <- subsetByOverlaps(txdb, true_TSS)
subset <- subsetByOverlaps(true_TSS, txdb, invert = TRUE)
promoters <- c("A" = length(data.frame(txdb)$name),
               "A&B" = length(data.frame(intersect)$name),
               "B" = length(data.frame(subset)$name))
promoters <- eulerr::euler(promoters)
plot(promoters, labels = c("Annotated\nPromoters", "CAGE\nPromoters"), 
     fills = list(fill = c("springgreen", "royalblue"), alpha = 0.33), quantities = TRUE)
```

To obtain the active promoters in breast samples, ctss from three samples (Normal, MCF7 and MDA-MB-453) were overlapped with annotated promoters and its complement subset of *true_TSS*. Additionally, promoter scores were calculated as the sum of overlapped ctss scores, and gene scores as the sum of ctss scores assigned to each gene.

```{r scores}
samples <- list(normal_ctss, MCF7_ctss, MDA_MB_453_ctss)
names(samples) <- c("Normal", "MCF7", "MDA-MB-453")
scores <- list()
for (i in 1:length(samples)) {
  total <- sum(samples[[i]]$score)
  data.frame(mcols(samples[[i]])) %>% 
    select(ctss = name, score) %>% 
    mutate(score = score / total) -> mcols(samples[[i]])
  
  mergeByOverlaps(txdb, samples[[i]]) %>% 
    data.frame() %>% 
    select(name, gene, ctss, score) %>% 
    distinct(ctss, .keep_all = TRUE) %>% 
    rbind(
      mergeByOverlaps(subset, samples[[i]]) %>% 
        data.frame() %>% 
        select(name, gene, ctss, score) %>% 
        distinct(ctss, .keep_all = TRUE)
    ) -> merge

  scores[[i]] <- merge %>%
    group_by(name) %>% mutate(promoter_score = sum(score)) %>% ungroup() %>% 
    group_by(gene) %>% mutate(gene_score = sum(score)) %>% ungroup() %>% 
    select(-c(ctss, score)) %>% 
    unique()
  names(scores)[i] <- names(samples)[i]
}
```

```{r results}
df <- data.frame()
for (i in 1:length(scores)) {
  tmp <- cbind(scores[[i]], sample = names(scores)[i])
  df <- rbind(df, tmp)
}
df <- mutate(df, sample = factor(sample, levels = names(scores)))
```

Promoters per gene were counted.

```{r promoters per gene}
df %>% 
  select(-c(promoter_score, gene_score, sample)) %>% 
  unique() %>% 
  group_by(gene) %>% 
  add_tally() %>% 
  ungroup() %>% 
  select(gene, n) %>% 
  unique() %>% 
  ggplot(aes(x = n)) +
  geom_histogram(stat = "count", color = "black", fill = "grey") +
  xlab("promoters per gene") +
  ylab("gene count") +
  ylim(c(0, 20000)) +
  theme_linedraw() +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(breaks = c(seq(1, 20, by = 1)), labels = c(seq(1, 19, by = 1), "20+"), limits = c(0, 21))
```

The ratio between promoter score and gene score was analyzed in order to filter promoters within genes.

```{r ratio}
df %>%
  select(-sample) %>%
  unique() %>%
  mutate(ratio = promoter_score / gene_score) %>%
  ggplot(aes(x = ratio)) +
  geom_histogram(bins = 50, color = "black", fill = "grey") +
  xlab(expression(over("promoter score", "gene score"))) +
  ylab("promoter count") +
  theme_linedraw() +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(n.breaks = 10)
```

With the goal of filtering promoters with fewer representation within genes, promoters were analyzed by a density plot.

```{r density_plot}
df %>% 
  select(-sample) %>% 
  unique() %>% 
  mutate(ratio = promoter_score / gene_score) %>% 
  filter(ratio < 0.5) %>% 
  ggplot(aes(x = ratio)) +
  geom_density() +
  xlab(expression(over("promoter score", "gene score"))) +
  theme_linedraw() +
  theme(panel.grid = element_blank()) -> density_plot
density_plot <- ggplot_build(density_plot)
density_plot$plot
```

From plot analysis, a cutoff was chosen.

```{r cutoff}
cutoff <- density_plot$data[[1]][which(diff(sign(diff(density_plot$data[[1]]$y))) == +2) + 1, ][2, ]
cutoff <- cutoff$x
```

A cutoff of `r signif(cutoff, 2)` was selected to filter promoters within genes by their relative score.

```{r density_plot with cutoff}
density_plot$plot + geom_vline(xintercept = cutoff, color = "red") +
  geom_ribbon(data = subset(density_plot$data[[1]], x > cutoff),
              aes(x = x, ymax = y, ymin = 0), fill = "blue", alpha = 0.5)
```

A empirical cumulative density function (ecdf) plot with the selected cutoff was plotted.

```{r ecdf_plot}
df %>% 
  select(-sample) %>% 
  unique() %>% 
  mutate(ratio = promoter_score / gene_score) %>% 
  filter(ratio < 0.5) %>% 
  ggplot(aes(x = ratio)) +
  stat_ecdf() +
  xlab(expression(over("promoter score", "gene score"))) +
  theme_linedraw() +
  theme(panel.grid = element_blank()) -> ecdf_plot
ecdf_plot <- ggplot_build(ecdf_plot)
ecdf_plot$plot + geom_vline(xintercept = cutoff, color = "red") +
  geom_ribbon(data = subset(ecdf_plot$data[[1]], x > cutoff),
              aes(x = x, ymax = y, ymin = 0), fill = "blue", alpha = 0.5)
```

```{r ratio with cutoff}
df %>% 
  select(-sample) %>% 
  unique() %>% 
  mutate(ratio = promoter_score / gene_score) %>% 
  ggplot(aes(x = ratio)) +
  geom_histogram(bins = 50, color = "black", fill = "grey") +
  geom_vline(xintercept = cutoff, color = "red") +
  xlab(expression(over("promoter score", "gene score"))) +
  ylab("promoter count") +
  theme_linedraw() +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(n.breaks = 10)
```

```{r filtered results}
df %>% 
  filter(promoter_score > gene_score * cutoff) -> df
```

Promoters per gene were counted after filtering.

```{r filtered promoters per gene}
df %>% 
  select(-c(promoter_score, gene_score, sample)) %>% 
  unique() %>% 
  group_by(gene) %>% 
  add_tally() %>% 
  ungroup() %>% 
  select(gene, n) %>% 
  unique() %>% 
  ggplot(aes(x = n)) +
  geom_histogram(stat = "count", color = "black", fill = "grey") +
  xlab("promoters per gene") +
  ylab("gene count") +
  theme_linedraw() +
  theme(panel.grid = element_blank()) +
  scale_x_continuous(breaks = c(seq(1, 8, by = 1)), labels = c(seq(1, 8, by = 1)), limits = c(0, 9))
```

The overlapping of promoter scores distributions can be interpreted as shared characteristics within analyzed samples.

```{r samples density_plot}
ggplot(df %>% filter(promoter_score > 0.0001), aes(log2(promoter_score), fill = sample)) + 
  geom_density(alpha = 0.33) + 
  scale_fill_manual(values = c("tomato", "springgreen", "royalblue")) + 
  xlab(expression(log[2]*"(promoter score > 0.0001)")) + 
  theme_linedraw() +
  theme(legend.position = "top", legend.title = element_blank(), panel.grid = element_blank())
```

The Venn diagram shows promoters across samples. Interestingly, almost half of active promoters are shared by the three analyzed samples.

```{r samples venn_diagram}
list <- list()
for (i in 1:length(scores)) {
  list[[i]] <- subset(scores[[i]]$name, scores[[i]]$name %in% df$name)
  names(list)[i] <- names(scores)[i]
}
list <- eulerr::euler(list)
plot(list, fills = list(fill = c("tomato", "springgreen", "royalblue"), alpha = 0.33),
     quantities = TRUE)
```

All active promoters were selected.

```{r selected promoters}
df %>% 
  filter(str_starts(name, "uc")) %>% 
  select(name, promoter_score, gene, gene_score, sample) %>% 
  left_join(data.frame(txdb)) %>% 
  rbind(
    df %>% 
      filter(str_starts(name, "p")) %>% 
      select(name, promoter_score, gene, gene_score, sample) %>% 
      left_join(data.frame(subset))
  ) -> results
```

Promoter regions were reduced to a width of 1.

```{r TSS}
results_TSS <- results
for (i in 1:nrow(results)) {
  if (rowMeans(cbind(results$start[i], results$end[i])) %% 1 == 0) {
    results_TSS$start[i] = results_TSS$end[i] = rowMeans(cbind(results$start[i], results$end[i]))
  } else {
    results_TSS$start[i] = results_TSS$end[i] = rowMeans(cbind(results$start[i], results$end[i])) - 0.5
  }
  results_TSS$width[i] = 1
}
```

```{r, include=FALSE}
results_TSS %>% 
  select(gene) %>%
  write_csv("tables/active_promoters.csv")
```

Finally, HGNC symbols were checked using the [multi-symbol checker](https://www.genenames.org/tools/multi-symbol-checker/) and promoter regions were extended 1000 nucleotides upstream and 200 nucleotides downstream from the TSS.

```{r HGNC symbols}
results_TSS %>% 
  left_join(read_csv("data/HGNC/active_promoters_hgnc-symbol-check.csv", skip = 1),
            by = c("gene" = "Input")) %>% 
  select(seqnames, start, end, width, strand,
         name, promoter_score, gene = `Approved symbol`, gene_score, sample) %>% 
  filter(!is.na(gene)) -> active_promoters_TSS
write_csv(active_promoters_TSS, "tables/active_promoters_TSS.csv")
active_promoters_TSS %>% 
  GRanges() %>% 
  promoters(upstream = 1000, downstream = 200) -> active_promoters
```

CDS retrieved from UCSC were subtracted from the extended promoter regions.

```{r cds}
cds <- cds(TxDb.Hsapiens.UCSC.hg19.knownGene)

subtract <- function(gr1, gr2) {
  # Subtract GRange object gr2 from gr1, but unlike setdiff, preserve individual ranges in gr1
  df1 = data.frame(seqnames = seqnames(gr1), start = start(gr1) - 1, end = end(gr1), strand = strand(gr1), mcols(gr1))
  df2 = data.frame(seqnames = seqnames(gr2), start = start(gr2) - 1, end = end(gr2), strand = strand(gr2), mcols(gr2))
  # -1 --> convert to base-0 start for bedtools
  res = bt.subtract(df1, df2)
  if (length(res) == 0) {
    # subtraction has left nothing remaining. Return empty GRanges obj.
    return(GRanges())
  } else {
    colnames(res) = colnames(df1)
    res$start = res$start + 1
    # reset to base-1 notation consistent with GRanges
    return (GRanges(res))
  }
}

active_promoters <- subtract(active_promoters, cds) %>% data.frame()
```

```{r active promoters}
write_csv(active_promoters, "tables/active_promoters.csv")

active_promoters %>% 
  select(-c(name, sample), name = gene) %>% 
  unique() %>% 
  GRanges() %>% 
  `seqlevelsStyle<-`("NCBI") %>% 
  export.bed("tables/active_promoters.bed")
```


