---
title: "Functional Enrichment Analysis"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath('..'))
knitr::opts_chunk$set(
	error = FALSE,
	message = FALSE,
	warning = FALSE)
```

Loading packages:

```{r packages}
library(tidyverse)
library(ggridges)
library(gridExtra)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(enrichplot)
library(ComplexHeatmap)
library(factoextra)
library(cluster)
library(kableExtra)
```

```{r functions, include=FALSE}
order <- base::order
select <- dplyr::select
rename <- dplyr::rename
arrange <- dplyr::arrange
group_by <- dplyr::group_by
```

Loading results:

```{r results}
regulonbrca <- read_tsv("tables/regulonbrca.tsv", col_types = "ccdd")

NES <- read_csv("tables/candidates_NES.csv")

FDR <- read_csv("tables/candidates_FDR.csv")

names = c("LNA_vs_NIC", "LM2_vs_NIC", "HM_vs_NIC", "LM2_vs_LNA", "HM_vs_LNA", "HM_vs_LM2")
LFC <- list()
for (i in 1:length(names)) {
  LFC[[i]] <- read_csv(paste0("tables/LFC_", names[i], ".csv")) %>% 
    column_to_rownames()
  names(LFC)[i] <- names[i]
}
LFC %>% 
  map(select, log2FoldChange) %>% 
  flatten() %>% 
  data.frame() %>% 
  `colnames<-`(names) %>% 
  mutate("MR" = rownames(LFC[[1]])) %>% 
  relocate(MR, .before = everything()) %>% 
  arrange(MR) %>% 
  filter(MR %in% NES$MR) -> LFC
```

Plotting ridgeplots:

```{r ridgeplots, fig.show="hold", out.width="50%"}
LFC %>%
  pivot_longer(cols = -MR, names_to = "line", values_to = "LFC") %>% 
  mutate(line = gsub("\\_", " ", line)) %>% 
  ggplot(aes(x = LFC, y = line, fill = ..x..)) +
  geom_density_ridges_gradient() +
  xlab(expression(log[2]~"fold change")) + ylab("") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  scale_y_discrete(limits = c("LNA vs NIC", "LM2 vs NIC", "HM vs NIC",
                              "LM2 vs LNA", "HM vs LNA", "HM vs LM2")) +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())

NES %>% 
  pivot_longer(cols = -MR, names_to = "line", values_to = "NES") %>% 
  mutate(line = gsub("\\_", " ", line)) %>% 
  ggplot(aes(x = NES, y = line, fill = ..x..)) +
  geom_density_ridges_gradient() +
  xlab("NES") + ylab("") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  scale_y_discrete(limits = c("LNA vs NIC", "LM2 vs NIC", "HM vs NIC",
                              "LM2 vs LNA", "HM vs LNA", "HM vs LM2")) +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
```

Plotting NES-LFC correlation:

```{r corr, fig.show="hold", out.width="50%"}
NES %>% 
  left_join(LFC, by = c("MR" = "MR")) -> NES_LFC

colors <- c("LNA vs NIC" = rainbow(6)[1], "LM2 vs NIC" = rainbow(6)[2],
            "HM vs NIC" = rainbow(6)[3], "LM2 vs LNA" = rainbow(6)[4],
            "HM vs LNA" = rainbow(6)[5], "HM vs LM2" = rainbow(6)[6])
ggplot(NES_LFC) +
  geom_point(aes(LNA_vs_NIC.x, LNA_vs_NIC.y, color = ifelse(abs(LNA_vs_NIC.y) < 1, "notDE", "LNA vs NIC"))) +
  geom_point(aes(LM2_vs_NIC.x, LM2_vs_NIC.y, color = ifelse(abs(LM2_vs_NIC.y) < 1, "notDE", "LM2 vs NIC"))) +
  geom_point(aes(HM_vs_NIC.x, HM_vs_NIC.y, color = ifelse(abs(HM_vs_NIC.y) < 1, "notDE", "HM vs NIC"))) +
  geom_point(aes(LM2_vs_LNA.x, LM2_vs_LNA.y, color = ifelse(abs(LM2_vs_LNA.y) < 1, "notDE", "LM2 vs LNA"))) +
  geom_point(aes(HM_vs_LNA.x, HM_vs_LNA.y, color = ifelse(abs(HM_vs_LNA.y) < 1, "notDE", "HM vs LNA"))) +
  geom_point(aes(HM_vs_LM2.x, HM_vs_LM2.y, color = ifelse(abs(HM_vs_LM2.y) < 1, "notDE", "HM vs LM2"))) +
  labs(x = "NES", y = expression(log[2]~"fold change"), color = "") +
  scale_color_manual(values = colors) +
  theme_linedraw() +
  theme(panel.grid = element_blank())

c1 <- ggplot(NES_LFC) +
  geom_point(aes(LNA_vs_NIC.x, LNA_vs_NIC.y), color = ifelse(abs(NES_LFC$LNA_vs_NIC.y) < 1, "gray50", rainbow(6)[1])) +
  labs(x = "NES", y = expression(log[2]~"fold change"), title = "LNA vs NIC") +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
c2 <- ggplot(NES_LFC) +
  geom_point(aes(LM2_vs_NIC.x, LM2_vs_NIC.y), color = ifelse(abs(NES_LFC$LM2_vs_NIC.y) < 1, "gray50", rainbow(6)[2])) +
  labs(x = "NES", y = expression(log[2]~"fold change"), title = "LM2 vs NIC") +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
c3 <- ggplot(NES_LFC) +
  geom_point(aes(HM_vs_NIC.x, HM_vs_NIC.y), color = ifelse(abs(NES_LFC$HM_vs_NIC.y) < 1, "gray50", rainbow(6)[3])) +
  labs(x = "NES", y = expression(log[2]~"fold change"), title = "HM vs NIC") +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
c4 <- ggplot(NES_LFC) +
  geom_point(aes(LM2_vs_LNA.x, LM2_vs_LNA.y), color = ifelse(abs(NES_LFC$LM2_vs_LNA.y) < 1, "gray50", rainbow(6)[4])) +
  labs(x = "NES", y = expression(log[2]~"fold change"), title = "LM2 vs LNA") +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
c5 <- ggplot(NES_LFC) +
  geom_point(aes(HM_vs_LNA.x, HM_vs_LNA.y), color = ifelse(abs(NES_LFC$HM_vs_LNA.y) < 1, "gray50", rainbow(6)[5])) +
  labs(x = "NES", y = expression(log[2]~"fold change"), title = "HM vs LNA") +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
c6 <- ggplot(NES_LFC) +
  geom_point(aes(HM_vs_LM2.x, HM_vs_LM2.y), color = ifelse(abs(NES_LFC$HM_vs_LM2.y) < 1, "gray50", rainbow(6)[6])) +
  labs(x = "NES", y = expression(log[2]~"fold change"), title = "HM vs LM2") +
  theme_linedraw() +
  theme(legend.position = "none", panel.grid = element_blank())
grid.arrange(c1, c2, c3, c4, c5, c6, ncol = 3)
```

# GSEA

```{r GSEA}
all_MR <- read_rds("tables/all_MR.rds")

gsea_res <- list()
for (i in 1:length(names)) {
  all_MR[[i]] %>%
  pull(NES) %>%
  `names<-`(
    all_MR[[i]] %>%
      pull(Regulon) %>%
      bitr(fromType = "SYMBOL", toType = "ENTREZID", OrgDb = "org.Hs.eg.db", drop = FALSE) %>%
      pull(ENTREZID)
  ) -> geneList
  gsea_res[[i]] <- gsePathway(geneList, seed = TRUE)
  names(gsea_res)[i] <- names[i]
  gsea_res[[i]] <- arrange(gsea_res[[i]], desc(NES))
}

write_rds(gsea_res, "tables/gsea_res.rds")
```

```{r NESplot}
NESplot <- function(x, n = 20, cutoff = 0.05, title = NULL) {
  x = x@result %>% arrange(desc(NES)) %>% {
    rbind(filter(., NES > 0) %>% {if (nrow(.) >= n/2) head(., n/2) else .},
          filter(., NES < 0) %>% {if (nrow(.) >= n/2) tail(., n/2) else .}
    )
  }
  plot = ggplot(x, aes(x = NES, y = factor(Description, levels = Description[order(NES)]))) +
    geom_bar(
      fill = ifelse(x$p.adjust < cutoff, 
                    ifelse(x$NES < 0, "cornflowerblue", "salmon"),
                    "gray50"),
      stat = "identity",
      show.legend = FALSE
    ) +
    # geom_text(
    #   label = x$setSize, 
    #   hjust = ifelse(x$NES < 0, 1.5, -0.5),
    #   size = 3
    # ) +
    labs(y = "", title = title) +
    theme_linedraw() # +
    # theme(panel.grid = element_blank()) 
  return(plot)
}
```

```{r, echo=FALSE}
gsea_res <- read_rds("tables/gsea_res.rds")

for (i in 1:6) {
  NESplot(gsea_res[[i]]) %>% print()
}
```

```{r}
merge <- data.frame()
for (i in 1:length(names)) {
  if (length(merge) == 0) {
    merge <- gsea_res[[i]]@result %>% 
      mutate(contrast = rep(names(gsea_res)[i], nrow(gsea_res[[i]]))) %>% 
      arrange(desc(NES)) %>% {
        rbind(filter(., NES > 0) %>% {if (nrow(.) >= 10) head(., 10) else .},
              filter(., NES < 0) %>% {if (nrow(.) >= 10) head(., 10) else .}
        )
      }
  } else {
    merge <- rbind(
      merge, gsea_res[[i]]@result %>% 
        mutate(contrast = rep(names(gsea_res)[i], nrow(gsea_res[[i]]))) %>% 
        arrange(desc(NES)) %>% {
          rbind(filter(., NES > 0) %>% {if (nrow(.) >= 10) head(., 10) else .},
                filter(., NES < 0) %>% {if (nrow(.) >= 10) head(., 10) else .}
          )
        }
    )
  }
}

svg("plots/GSEA.svg")
merge %>% 
  pivot_wider(id_cols = contrast, names_from = Description, values_from = NES) %>% 
  mutate(contrast = gsub("\\_", " ", contrast)) %>% 
  column_to_rownames("contrast") %>% 
  data.matrix() %>% 
  Heatmap(
  name = "NES",
  row_names_side = "left",
  column_names_side = "top",
  # column_names_rot = 45,
  column_names_max_height = unit(15, "cm"),
  column_names_gp = grid::gpar(fontsize = 6),
  border = TRUE,
  rect_gp = grid::gpar(col = "black", lwd = 1),
  cluster_columns = FALSE,
  cluster_rows = FALSE,
  col = circlize::colorRamp2(c(-2, 0, 4), c("blue", "white", "red"))
  )
dev.off()
```

```{r GSEAplot, echo=FALSE, eval=FALSE, include=FALSE}
gsea_res <- read_rds("tables/gsea_res.rds")

for (i in 1:6) {
  gseaplot2(gsea_res[[i]], gsea_res[[i]]@result[1, 1], gsea_res[[i]]@result[1, 2]) %>% print()
}
```

# Clustering

```{r k}
NES_clusters <- NES %>%
  column_to_rownames("MR") %>%
  scale()

dist <- get_dist(NES_clusters, method = "spearman")

fviz_dist(dist, show_labels = FALSE, gradient = list(low = "blue", mid = "white", high = "red")) +
  labs(fill = expression(rho))

fviz_nbclust(NES_clusters, cluster::pam, k.max = 25, method = "wss", diss = dist) +
  labs(title = "") +
  xlab("number of clusters (k)") +
  ylab("total within sum of squares") +
  theme_linedraw() +
  theme(panel.grid = element_blank())

fviz_nbclust(NES_clusters, cluster::pam, k.max = 25, method = "silhouette", diss = dist) +
  labs(title = "") +
  xlab("number of clusters (k)") +
  ylab("average silhouette width") +
  theme_linedraw() +
  theme(panel.grid = element_blank())

fviz_nbclust(NES_clusters, cluster::pam, k.max = 25, method = "gap_stat", diss = dist) +
  labs(title = "") +
  xlab("number of clusters (k)") +
  ylab("gap statistic (k)") +
  theme_linedraw() +
  theme(panel.grid = element_blank())
```

```{r}
k = 4
```

```{r PAM}
NES_PAM <- pam(dist, k, diss = TRUE, nstart = 50)

NES_PAM$data <- NES_clusters

fviz_cluster(NES_PAM, geom = "point", axes = c(1, 2), show.clust.cent = FALSE, shape = "circle",
             main = "", xlab = "PC1 (66.4 %)", ylab = "PC2 (21.9 %)") +
  labs(color = "MRC", fill = "MRC") +
  theme_linedraw() +
  theme(panel.grid = element_blank())
```

```{r 3D}
# write_rds(NES_PAM, "tables/NES_PAM.rds")
# NES_PAM <- read_rds("tables/NES_PAM.rds")

prcomp <- prcomp(NES_PAM$data, rank. = 3)
components <- prcomp[["x"]]
components <- data.frame(components)
components$PC2 <- -components$PC2
components$PC3 <- -components$PC3
components$MRC <- as.character(NES_PAM$clustering)

plotly::plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3, color = ~MRC, colors = viridis::viridis(4)) %>% 
  plotly::add_markers(size = 12) %>% 
  plotly::layout(legend = list(title = list(text = '<b> MRC </b>')))
```

```{r MRC}
NES_clusters <- NES %>%
  mutate(MRC = NES_PAM$clustering) %>% 
  group_by(MRC) %>% 
  {aggregate(.[, 2:7], list(.$MRC), mean)} %>% 
  ungroup() %>% 
  rename(MRC = Group.1)

NES_clusters %>% 
  column_to_rownames("MRC") %>%  
  data.matrix() %>% 
  `colnames<-`(gsub("\\_", " ", colnames(.))) %>% 
  Heatmap(
  name = "NES",
  row_title = "MRC",
  row_names_side = "left",
  # show_row_names = FALSE,
  border = TRUE,
  rect_gp = grid::gpar(col = "black", lwd = 1),
  cluster_columns = FALSE,
  cluster_rows = FALSE,
  col = circlize::colorRamp2(c(-5, 0, 5), c("blue", "white", "red"))
  )

NES_PAM$clustering %>% 
  data.frame("MRC" = .) %>% 
  rownames_to_column("MR") -> MRC
write_csv(MRC, "tables/MRC_list.csv")

for (i in 1:k) {
  MRC %>% 
    filter(MRC == i) %>% 
    select(MR) %>% 
    write_csv(paste0("STRING/MRC", i, ".txt"), col_names = FALSE)
}
```

# STRING

With four sources of interaction information (experiments, databases, co-expression and gene fusion), a filter of highest confidence (0.900) of interaction, and the complete list of MR as universe, PPI and functional enrichment was assessed.

Out of 354 nodes (MR candidates), 361 edges (interactions) of 263 expected were found. PPI enrichment p-value of 5.77E-9.

```{r, echo=FALSE}
data.frame(
  "MRC" = c(1:4),
  "nodes" = c(81, 89, 75, 109),
  "edges" = c(5, 34, 17, 128),
  "expected" = c(3, 23, 5, 51),
  "PPI enrichment p-value" = c(0.173, 0.0227, "3.91E-5", "<1E-16"),
  "functional enrichment" = c(FALSE, TRUE, TRUE, TRUE),
  check.names = FALSE,
  fix.empty.names = FALSE
) %>% 
  kbl(align = "llllll") %>% 
  kable_classic("striped", full_width = FALSE, html_font = "Arial") %>% 
  row_spec(0, bold = TRUE)
```


